Gaussian Process regression for astronomical time-series

### 2022-08-08 src/tex/ms.tex
[showyourwork] overleaf sync

\begin{marginnote}[]
  \entry{GP}{Gaussian Process}
  \entry{GPR}{Gaussian Process Regression}
\end{marginnote}

@@

\begin{marginnote}[]
  \entry{Kernel function}{The ``kernel'' or ``covariance'' function, typeset here as $k(t_i,\,t_j;\,\phi)$ is a parametric description of the covariance between two data points $t_i$ and $t_j$. The parameter $\phi$ indicates that these functions are typically parameterized by a set of parameters, typically referred to as ``hyperparameters''.}
\end{marginnote}

\begin{marginnote}[]
  \entry{Hyperparameters}{In a GP model, the kernel function is typically parameterized by a set of parameters that we will label as $\phi$ and refer to as ``hyperparameters''. This label is used to distinguish these parameters from the parameters of the mean function, which we will refer to as $\theta$.}
\end{marginnote}

@@

To address this, \citet{2009ApJ...704...51C} proposed a wavelet-based method to model correlated noise in transit light curves, which can be seen as a special case of GPR with a covariance function belonging to the exponential family, but the formulation using wavelets. Rather than modelling the covariance in the time domain, the Power Spectral Density (PSD) of the noise is assumed to be of the form $P(f) \propto f^{-\gamma}$. 
Noting that the wavelet transform of such a process gives rise to a nearly diagonal covariance matrix, \citet{2009ApJ...704...51C}  derived an expression for the likelihood that can be computed in $\mathcal{O}(N)$ operations. This method has been used widely since its publication to model transit observations from space missions which stare continuously at a given field, such as \textit{Kepler} or \textit{TESS}. However, the use of a wavelet transform requires regular time sampling, which precludes its application to datasets with irregular sampling or significant data gaps, for example from space telescopes in low-Earth orbit such as Hubble and Spitzer. 
\begin{marginnote}
  \entry{Colours of noise}{White, pink and red noise correspond to $\gamma=0$, $1$ and $2$, respectively.}
\end{marginnote}

@@

One general shortcoming of current GP-based systematics models with multi-dimensional inputs is that the uncertainties on the input variables are ignored. This could in principle be remedied by treating the housekeeping variables as noisy observations of latent GP variables, and modelling them alongside the observed fluxes, but we are not aware of published attempts to do this in practice to date. \begin{marginnote}
  \entry{Latent variable}{A random variable which is never actually observed.}
\end{marginnote}
In spectroscopic observations, whether using GPs or parametric models, it has become standard practice to model the ``white" light curve first (obtained by integrating the spectrum over the full wavelength range at each time-step). ``Coloured'' light curves, extracted in individual wavelength bins, are then divided by the best-fit systematics model derived from the white light curve, before being modelled further. To our knowledge, no attempt has been made to model the wavelength dependence of the systematics directly.

@@

% \subsection{Sidebars and Margin Notes}
% % Margin Note
% \begin{marginnote}[]
%   \entry{Term A}{definition}
%   \entry{Term B}{definition}
%   \entry{Term C}{defintion}
% \end{marginnote}

>>>>

\begin{aaramarginnote}[]
  \entry{GP}{Gaussian Process}
  \entry{GPR}{Gaussian Process Regression}
\end{aaramarginnote}

@@

\begin{aaramarginnote}[]
  \entry{Kernel function}{The ``kernel'' or ``covariance'' function, typeset here as $k(t_i,\,t_j;\,\phi)$ is a parametric description of the covariance between two data points $t_i$ and $t_j$. The parameter $\phi$ indicates that these functions are typically parameterized by a set of parameters, typically referred to as ``hyperparameters''.}
\end{aaramarginnote}

\begin{aaramarginnote}[]
  \entry{Hyperparameters}{In a GP model, the kernel function is typically parameterized by a set of parameters that we will label as $\phi$ and refer to as ``hyperparameters''. This label is used to distinguish these parameters from the parameters of the mean function, which we will refer to as $\theta$.}
\end{aaramarginnote}

@@

To address this, \citet{2009ApJ...704...51C} proposed a wavelet-based method to model correlated noise in transit light curves, which can be seen as a special case of GPR with a covariance function belonging to the exponential family, but the formulation using wavelets. Rather than modelling the covariance in the time domain, the Power Spectral Density (PSD) of the noise is assumed to be of the form $P(f) \propto f^{-\gamma}$. 
Noting that the wavelet transform of such a process gives rise to a nearly diagonal covariance matrix, \citet{2009ApJ...704...51C}  derived an expression for the likelihood that can be computed in $\mathcal{O}(N)$ operations. This method has been used widely since its publication to model transit observations from space missions which stare continuously at a given field, such as \textit{Kepler} or \textit{TESS}. However, the use of a wavelet transform requires regular time sampling, which precludes its application to datasets with irregular sampling or significant data gaps, for example from space telescopes in low-Earth orbit such as Hubble and Spitzer. 
\begin{aaramarginnote}
  \entry{Colours of noise}{White, pink and red noise correspond to $\gamma=0$, $1$ and $2$, respectively.}
\end{aaramarginnote}

@@

One general shortcoming of current GP-based systematics models with multi-dimensional inputs is that the uncertainties on the input variables are ignored. This could in principle be remedied by treating the housekeeping variables as noisy observations of latent GP variables, and modelling them alongside the observed fluxes, but we are not aware of published attempts to do this in practice to date. \begin{aaramarginnote}
  \entry{Latent variable}{A random variable which is never actually observed.}
\end{aaramarginnote}
In spectroscopic observations, whether using GPs or parametric models, it has become standard practice to model the ``white" light curve first (obtained by integrating the spectrum over the full wavelength range at each time-step). ``Coloured'' light curves, extracted in individual wavelength bins, are then divided by the best-fit systematics model derived from the white light curve, before being modelled further. To our knowledge, no attempt has been made to model the wavelength dependence of the systematics directly.

@@

% \subsection{Sidebars and Margin Notes}
% % Margin Note
% \begin{aaramarginnote}[]
%   \entry{Term A}{definition}
%   \entry{Term B}{definition}
%   \entry{Term C}{defintion}
% \end{aaramarginnote}
